---
title: "PHEM KPI Dashboard"
logo: ../assets/ephi_logo.png
format:
  dashboard:
    theme: [flatly, dashboard.scss]
    embed-resources: true
    nav-buttons:
      - icon: globe
        href: https://www.ephi.gov.et
    orientation: rows
    scrolling: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr); library(readr); library(tidyr); library(stringr)
library(plotly); library(leaflet); library(DT); library(sf)

# Quarto renders with the working directory set to the .qmd folder (reports/).
root <- normalizePath(file.path(getwd(), ".."))
op   <- function(...) file.path(root, "output", ...)
tb   <- function(...) file.path(root, "exports", "tables", ...)

source(file.path(root, "scripts", "catalog.R"))   # PERIOD_LEVELS/LABELS, FOCUS_PERIOD

national <- read_csv(op("kpi_national.csv"),       show_col_types = FALSE)
region   <- read_csv(op("kpi_region.csv"),         show_col_types = FALSE)
wide     <- read_csv(op("master_woreda_wide.csv"), show_col_types = FALSE)
wpcode   <- read_csv(op("kpi_woreda_pcode.csv"),   show_col_types = FALSE)
catalog  <- read_csv(op("indicator_catalog.csv"),  show_col_types = FALSE)

trend_nat  <- read_csv(tb("trend_national_by_period.csv"), show_col_types = FALSE)
trend_comp <- read_csv(tb("trend_region_composite.csv"),  show_col_types = FALSE)
trend_dq   <- read_csv(tb("trend_data_quality.csv"),      show_col_types = FALSE)
trend_cf   <- read_csv(tb("trend_corefunction.csv"),      show_col_types = FALSE)
cov        <- read_csv(tb("geographic_coverage.csv"),     show_col_types = FALSE)

# Reusable woreda geometry from the GIS shapefile (1,082 woredas keyed by
# adm3_pcode). It joins directly to kpi_woreda_pcode.csv. See assets/README.
geo3 <- readRDS(file.path(root, "assets", "eth_adm3_geo.rds"))
geo1 <- readRDS(file.path(root, "assets", "eth_adm1_geo.rds"))

# --- GUARDRAIL: proportions must never exceed 100% ---------------------------
# The pipeline already caps and validates pairs; assert it here so a bad render
# fails loudly rather than shipping a >100% figure.
bad <- bind_rows(national, region) |>
  filter(value_type == "proportion", !is.na(value_pct), value_pct > 100.0001)
if (nrow(bad) > 0) stop("Guardrail tripped: ", nrow(bad), " proportion(s) > 100%.")

# --- conveniences ------------------------------------------------------------
FP        <- FOCUS_PERIOD
FP_LABEL  <- unname(PERIOD_LABELS[FP])
prop_only <- function(df) filter(df, value_type == "proportion")
pf        <- function(p) factor(p, levels = PERIOD_LEVELS,
                                labels = unname(PERIOD_LABELS[PERIOD_LEVELS]))
EPHI_GREEN <- "#006600"

# performance palette: red (poor) -> green (good), per house style
perf_pal <- function(domain = c(0, 100))
  colorNumeric("RdYlGn", domain = domain, na.color = "#e6e6e6")
RYG <- list(c(0, "#d73027"), c(0.5, "#fee08b"), c(1, "#1a9850"))

# focus-period slices
nat_fp <- national |> filter(period == FP)
reg_fp <- region   |> filter(period == FP)
wide_fp <- wide    |> filter(period == FP)

# reporting level per region (woreda / facility), from the master. Used to label
# regions that did NOT report at woreda level so coverage is not over-claimed.
region_level <- read_csv(op("master_woreda.csv"),
                         col_select = c(region, period, report_level),
                         show_col_types = FALSE) |>
  filter(period == FP) |> distinct(region, report_level)
```

```{r headline-numbers}
N_REGIONS_TOTAL <- 14L     # regions / city administrations in the 2021 reference
N_WOREDAS_REF   <- 1082L   # reference woredas in reference_geo_names.csv

n_reg <- n_distinct(reg_fp$region)
n_wor <- wide_fp |> distinct(region, woreda) |> nrow()
coverage_nat <- round(100 * n_wor / N_WOREDAS_REF)

entries_filled <- trend_dq |> filter(period == FP) |> pull(pct_filled)
entries_filled <- if (length(entries_filled)) round(entries_filled) else NA_real_

nat_prop  <- prop_only(nat_fp)

cov_col <- function(x) if (isTRUE(x >= 66)) "success" else
                       if (isTRUE(x >= 33)) "warning" else "danger"

# Per-region reporting + performance, so a high % is always read next to how
# much was actually reported (cautious interpretation).
reg_perf <- reg_fp |>
  prop_only() |>
  group_by(region) |>
  summarise(`indicators (of 21)`  = sum(!is.na(value_pct_capped)),
            `mean performance %`   = round(mean(value_pct_capped, na.rm = TRUE)),
            .groups = "drop")
reg_units <- wide_fp |> group_by(region) |>
  summarise(`units reported` = n_distinct(woreda), .groups = "drop")
reg_summary <- reg_units |>
  left_join(region_level, by = "region") |>
  left_join(cov |> select(region, `% woreda coverage` = coverage_pct), by = "region") |>
  left_join(reg_perf, by = "region") |>
  # label the reporting level so woreda coverage is not over-claimed:
  #   facility  -> Dire Dawa reports per health facility (PHCU/hospital)
  #   aggregate -> a single region-level row (Harari)
  # woreda coverage only applies to woreda-level regions; Dire Dawa (facility)
  # and Harari (single aggregate row) are blanked (see note under the table).
  mutate(.is_woreda = report_level != "facility" & `units reported` > 1,
         `% woreda coverage` = ifelse(.is_woreda, `% woreda coverage`, NA_real_)) |>
  select(region, `units reported`, `% woreda coverage`,
         `indicators (of 21)`, `mean performance %`) |>
  arrange(desc(`units reported`))

# National mean by PHEM core function (current period) for the reporting page.
cf_fp <- nat_prop |>
  group_by(group) |>
  summarise(v = round(mean(value_pct_capped, na.rm = TRUE)),
            n = sum(!is.na(value_pct_capped)), .groups = "drop") |>
  arrange(v)
```

# Overview

## {.toolbar}

**Reporting period: `r FP_LABEL`.** Performance figures reflect woredas that reported.

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Regions reporting"
list(icon = "buildings", color = "primary",
     value = paste0(n_reg, " of ", N_REGIONS_TOTAL))
```

```{r}
#| content: valuebox
#| title: "Woredas reporting"
list(icon = "geo-alt", color = "info", value = format(n_wor, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Woreda coverage (of 2021 reference)"
list(icon = "pin-map", color = cov_col(coverage_nat),
     value = paste0(coverage_nat, "%"))
```

```{r}
#| content: valuebox
#| title: "KPI cells filled (data completeness)"
list(icon = "ui-checks", color = cov_col(entries_filled),
     value = paste0(entries_filled, "%"))
```

## Row

::: {.callout-note collapse="true" title="Disclaimer/Precaution"}
All KPI percentages are computed **among woredas that reported** (summed
numerator ÷ summed denominator), so they describe performance among reporters,
not across all woredas. Read every performance figure next to the reporting
columns: a high percentage built on a handful of reporting woredas (e.g. Oromia
this period) is not comparable to one built on most of a region. Coverage is
measured against the 2021 reference list, which undercounts today's woredas, so
it is an upper bound. Three regions did not report at all this period.
:::

## Row

### Column {width="48%"}

```{r}
#| title: "Regional reporting & performance — read together"
DT::datatable(reg_summary, rownames = FALSE,
              options = list(dom = "t", pageLength = 20, scrollX = TRUE)) |>
  DT::formatStyle("units reported",
                  background = DT::styleColorBar(range(reg_summary$`units reported`),
                                                 "#cfe8cf"),
                  backgroundSize = "98% 70%", backgroundRepeat = "no-repeat",
                  backgroundPosition = "center")
```

::: {.callout-tip collapse="true" title="Column definitions"}
- **units reported** — distinct reporting units for the region (woredas in most regions).
- **% woreda coverage** — reporting woredas ÷ that region's 2021 reference woredas (an upper bound; woreda-level only).
- **indicators (of 21)** — how many of the 21 proportion KPIs produced a usable value this period. The rest were either left blank or had no events to measure (a zero denominator, e.g. no outbreaks), so they are N/A.
- **mean performance %** — simple average of the region's usable KPI percentages (each = Σ numerator ÷ Σ denominator among reporting woredas, capped at 100). A summary only — always read it with the reporting columns to its left.
- **NB:** Dire Dawa reports per health facility (PHCU/hospital) and Harari as a single regional row, so their *units reported* are not woredas and woreda coverage is not shown for them.
:::

### Column {width="52%"}

```{r}
#| title: "National KPI performance — among woredas that reported"
d <- nat_prop |>
  mutate(label = str_trunc(label, 52)) |>
  arrange(value_pct_capped)
d$label <- factor(d$label, levels = d$label)
plot_ly(d, x = ~value_pct_capped, y = ~label, type = "bar",
        orientation = "h",
        marker = list(color = ~value_pct_capped, colorscale = RYG,
                      cmin = 0, cmax = 100),
        hovertemplate = "%{y}<br>%{x:.0f}%<extra></extra>") |>
  layout(xaxis = list(title = "% (capped at 100; dashed line = 80% target)",
                      range = c(0, 105)),
         yaxis = list(title = ""),
         shapes = list(list(type = "line", x0 = 80, x1 = 80,
                            y0 = -0.5, y1 = nrow(d) - 0.5,
                            line = list(color = "grey40", dash = "dash"))),
         margin = list(l = 320))
```

## Row {height="30%"}

```{r}
#| title: "Performance by PHEM core function — national (mean of indicators)"
cf_fp$group <- factor(cf_fp$group, levels = cf_fp$group)
plot_ly(cf_fp, x = ~v, y = ~group, type = "bar", orientation = "h",
        marker = list(color = ~v, colorscale = RYG, cmin = 0, cmax = 100),
        text = ~paste0(n, " indicators"),
        hovertemplate = "%{y}<br>%{x:.0f}% (mean of %{text})<extra></extra>") |>
  layout(xaxis = list(title = "mean % of indicators (capped at 100)",
                      range = c(0, 105)),
         yaxis = list(title = ""))
```

# Regional

## {.toolbar}

**Period: `r FP_LABEL`.**

## Row

```{r}
#| title: "Region × indicator performance (%)"
mat <- reg_fp |>
  prop_only() |>
  mutate(label = str_trunc(label, 46)) |>
  select(region, label, value_pct_capped)

natcol <- nat_prop |>
  mutate(label = str_trunc(label, 46), region = "NATIONAL") |>
  select(region, label, value_pct_capped)

hm <- bind_rows(mat, natcol) |>
  pivot_wider(names_from = region, values_from = value_pct_capped)

# order columns with NATIONAL last, order rows by national performance
ord <- natcol |> arrange(desc(value_pct_capped)) |> pull(label)
hm  <- hm[match(ord, hm$label), ]
labs <- hm$label
M    <- as.matrix(hm[, -1])
cols <- c(setdiff(colnames(M), "NATIONAL"), "NATIONAL")
M    <- M[, cols]

plot_ly(x = colnames(M), y = labs, z = M, type = "heatmap",
        colorscale = list(c(0, "#d73027"), c(0.5, "#fee08b"), c(1, "#1a9850")),
        zmin = 0, zmax = 100,
        hovertemplate = "%{x}<br>%{y}<br>%{z:.0f}%<extra></extra>") |>
  layout(xaxis = list(title = "", tickangle = -45),
         yaxis = list(title = "", automargin = TRUE),
         margin = list(l = 300, b = 120))
```

## Row {height="40%"}

### Column

```{r}
#| title: "Regional composite score"
comp <- trend_comp |>
  filter(period == FP) |>
  arrange(composite)
comp$region <- factor(comp$region, levels = comp$region)
plot_ly(comp, x = ~composite, y = ~region, type = "bar", orientation = "h",
        marker = list(color = ~composite,
                      colorscale = list(c(0, "#d73027"), c(0.5, "#fee08b"),
                                        c(1, "#1a9850")),
                      cmin = 0, cmax = 100),
        hovertemplate = "%{y}<br>composite %{x:.0f}%<extra></extra>") |>
  layout(xaxis = list(title = "Composite (mean of proportions, %)",
                      range = c(0, 100)),
         yaxis = list(title = ""))
```

### Column

```{r}
#| title: "Region table"
reg_tab <- reg_fp |>
  prop_only() |>
  group_by(region) |>
  summarise(
    `indicators (of 21)` = sum(!is.na(value_pct_capped)),
    `mean performance %` = round(mean(value_pct_capped, na.rm = TRUE)),
    .groups = "drop") |>
  arrange(desc(`mean performance %`))
datatable(reg_tab, rownames = FALSE, options = list(dom = "t", pageLength = 20))
```

# Trends

## {.toolbar}

**Periods: `r paste(unname(PERIOD_LABELS[PERIOD_LEVELS]), collapse = "  ·  ")`.**

::: {.callout-note title="Comparing periods of different length"}
`r FP_LABEL` (weeks 28–52) spans **two quarters**, while
`r paste(unname(PERIOD_LABELS[head(PERIOD_LEVELS, -1)]), collapse = " and ")`
are single quarters. Proportion KPIs are ratios, so they remain comparable across
periods. Only raw counts scale with period length.
:::

## Row

### Column {width="58%"}

```{r}
#| title: "National KPI trajectory — indicator × period (read each row left→right)"
# A heatmap instead of 21 overlapping lines: each row is one indicator, columns
# are the periods, colour is the value. Trajectory = how a row's colour shifts.
tn  <- trend_nat |> mutate(label = str_trunc(label, 44), period_lab = pf(period))
ord <- tn |> group_by(label) |>
  summarise(m = mean(value_pct, na.rm = TRUE), .groups = "drop") |>
  arrange(m) |> pull(label)
W <- tn |> select(label, period_lab, value_pct) |>
  pivot_wider(names_from = period_lab, values_from = value_pct)
W <- W[match(ord, W$label), ]
labs <- W$label; Z <- as.matrix(W[, -1])
plot_ly(x = colnames(Z), y = labs, z = Z, type = "heatmap",
        colorscale = RYG, zmin = 0, zmax = 100,
        hovertemplate = "%{y}<br>%{x}: %{z:.0f}%<extra></extra>") |>
  layout(xaxis = list(title = ""), yaxis = list(title = "", automargin = TRUE),
         margin = list(l = 290))
```

### Column {width="42%"}

```{r}
#| title: "PHEM core functions over time"
cf <- trend_cf |> mutate(period_lab = pf(period))
plot_ly(cf, x = ~period_lab, y = ~value, color = ~group,
        type = "scatter", mode = "lines+markers",
        hovertemplate = "%{fullData.name}<br>%{x}: %{y:.0f}%<extra></extra>") |>
  layout(xaxis = list(title = ""),
         yaxis = list(title = "mean %", range = c(0, 105)),
         legend = list(font = list(size = 10)))
```

## Row {height="42%"}

```{r}
#| title: "Regional composite — change from earliest to latest period reported"
# A dumbbell: grey dot = earliest period, coloured dot = latest (green if it rose,
# red if it fell). Regions that reported only once show a single dot.
po <- setNames(seq_along(PERIOD_LEVELS), PERIOD_LEVELS)
dc <- trend_comp |>
  mutate(o = po[period]) |>
  group_by(region) |>
  summarise(first = composite[which.min(o)], last = composite[which.max(o)],
            .groups = "drop") |>
  mutate(dir = ifelse(last >= first, "improved", "declined")) |>
  arrange(last)
dc$region <- factor(dc$region, levels = dc$region)
plot_ly(dc) |>
  add_segments(x = ~first, xend = ~last, y = ~region, yend = ~region,
               line = list(color = "grey75", width = 3),
               showlegend = FALSE, hoverinfo = "skip") |>
  add_markers(x = ~first, y = ~region, name = "earliest",
              marker = list(color = "grey55", size = 9),
              hovertemplate = "earliest: %{x:.0f}%<extra></extra>") |>
  add_markers(x = ~last, y = ~region, name = "latest",
              marker = list(color = ~ifelse(dir == "improved", "#1a9850", "#d73027"),
                            size = 12),
              hovertemplate = "latest: %{x:.0f}%<extra></extra>") |>
  layout(xaxis = list(title = "composite % (earliest → latest period reported)",
                      range = c(0, 100)),
         yaxis = list(title = ""),
         legend = list(orientation = "h", y = -0.18))
```

# Map

## {.toolbar}

**Period: `r FP_LABEL`.**

## Row

::: {.card title="Woreda performance map"}

```{r}
# Build a leaflet choropleth for a chosen indicator (or composite) over the
# focus period, joining woreda rollups to the GIS geometry on adm3_pcode.
make_map <- function(indicator_code = NULL, title = "") {
  if (is.null(indicator_code)) {
    vals <- wpcode |>
      filter(period == FP, value_type == "proportion") |>
      group_by(adm3_pcode) |>
      summarise(v = mean(value_pct_capped, na.rm = TRUE), .groups = "drop")
  } else {
    vals <- wpcode |>
      filter(period == FP, indicator == indicator_code) |>
      group_by(adm3_pcode) |>
      summarise(v = mean(value_pct_capped, na.rm = TRUE), .groups = "drop")
  }
  g <- geo3 |> left_join(vals, by = "adm3_pcode")
  pal <- perf_pal()
  lab <- sprintf("<b>%s</b><br/>%s: %s",
                 g$adm3_en, title,
                 ifelse(is.na(g$v), "no data", paste0(round(g$v), "%"))) |>
    lapply(htmltools::HTML)
  leaflet(g, options = leafletOptions(minZoom = 5)) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addPolygons(fillColor = ~pal(v), fillOpacity = 0.85, weight = 0.4,
                color = "white", label = lab,
                highlightOptions = highlightOptions(weight = 2, color = "#333",
                                                    bringToFront = TRUE)) |>
    addPolylines(data = geo1, color = "#444", weight = 1, opacity = 0.7) |>
    addLegend(pal = pal, values = c(0, 100), title = paste0(title, " (%)"),
              position = "bottomright") |>
    setView(lng = 39.5, lat = 9.0, zoom = 6)
}
```

::: {.panel-tabset}

## Composite

```{r}
make_map(NULL, "Composite")
```

## Timeliness

```{r}
make_map("RPT_TIMELY", "Reports on time")
```

## Completeness

```{r}
make_map("RPT_COMPLETE", "Reports complete")
```

## Response 24–48h

```{r}
make_map("RESP_2448", "Response 24–48h")
```

:::

:::

# Data quality

## {.toolbar}

**Metrics table covers all periods; the completeness chart is for `r FP_LABEL`.**

## Row {height="35%"}

```{r}
#| title: "Data-quality metrics by period"
dq <- trend_dq |>
  mutate(period = unname(PERIOD_LABELS[period])) |>
  rename(`regions` = regions_reporting, `woreda units` = woreda_units,
         `mapped woredas` = mapped_woredas, `records` = records,
         `% cells filled` = pct_filled,
         `num > denom (errors)` = impossible_num_gt_denom,
         `raw > 100% (pre-cap)` = over_100_raw)
datatable(dq, rownames = FALSE, options = list(dom = "t"))
```

## Row

```{r}
#| title: "Reporting completeness — % of woredas reporting each indicator"
rc <- nat_fp |>
  mutate(label = str_trunc(label, 52)) |>
  arrange(pct_reported)
rc$label <- factor(rc$label, levels = rc$label)
plot_ly(rc, x = ~pct_reported, y = ~label, type = "bar", orientation = "h",
        marker = list(color = ~pct_reported,
                      colorscale = list(c(0, "#d73027"), c(0.5, "#fee08b"),
                                        c(1, "#1a9850")),
                      cmin = 0, cmax = 100),
        hovertemplate = "%{y}<br>%{x:.0f}% of woredas reported<extra></extra>") |>
  layout(xaxis = list(title = "% of woredas reporting", range = c(0, 105)),
         yaxis = list(title = ""), margin = list(l = 320))
```

# Data

## {.toolbar}

**Period: `r FP_LABEL`.**

## Row

```{r}
#| title: "Woreda × indicator values — search and export"
dt <- wide_fp |>
  select(region, zone, woreda, population_best,
         RPT_TIMELY, RPT_COMPLETE, NOTIFY_24H, RESP_2448, LAB_7D,
         EPRP_PLAN, DHIS2_RPT, ATTACK_RATE, CFR) |>
  mutate(across(where(is.numeric), ~round(.x, 1)))
datatable(dt, rownames = FALSE, filter = "top",
          extensions = "Buttons",
          options = list(pageLength = 25, scrollX = TRUE,
                         dom = "Bfrtip", buttons = c("csv", "excel")))
```
