---
title: "PHEM KPI Pipeline — Developer & Reproducibility Guide"
subtitle: "The approach in concept, then a teaching walkthrough of the code"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    code-fold: false
    code-overflow: wrap
  docx:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr); library(dplyr); library(knitr)
root <- normalizePath(file.path(getwd(), ".."))
op <- function(...) file.path(root, "output", ...)
```

# Part I — The approach, in concept

This first part explains *how the pipeline thinks* before any code. Read it once
and the scripts in Part II will read as an implementation of these ideas rather
than a wall of R.

## 1.1 The problem

Each region submits a PHEM KPI workbook. The workbooks describe the same 23
indicators, but they are not standardised. Sheet names differ, a title row sits
above the header in some files, one region carries a Population column and others
do not, Addis reports by sub-city, Dire Dawa reports by health facility, and a
proportion is written on a 0–1 scale in one file and 0–100 in another. We now
hold **three reporting periods** of these workbooks, and more will arrive.

The goal is a single, tidy, reproducible dataset from which regional and national
performance, maps, and a written report follow automatically, and which extends
to each new quarter with a small, documented change.

## 1.2 The central idea: one long tidy table keyed by period

Everything rests on one table, `output/master_woreda.csv`, with exactly one row
per **woreda × indicator × period**. This "long" grain is the design decision
that makes the rest simple.

```
region  period        woreda    indicator    numerator  denominator  value_reported
Afar    2025-Wk28-52  Adaar     RPT_TIMELY    18         20           90
Afar    2025-Wk28-52  Adaar     RESP_2448     1          1            100
Afar    2025-Wk16-27  Adaar     RPT_TIMELY    9          12           75
...
```

Because `period` is a column rather than a separate file, adding a quarter does
not change the schema. Region rollups, national rollups, the woreda map, and the
cross-period trends are all *the same table grouped differently*. A new period
simply adds rows.

## 1.3 The four rules that make numbers trustworthy

These are the analytic commitments. They matter more than any single line of
code, and they are the reason the figures can be defended.

**Rule 1 — Recompute every proportion from counts.** A reported percentage cannot
be trusted (0–1 here, 0–100 there, rounded elsewhere). We keep the numerator and
denominator and compute the percentage ourselves:

$$\text{value} = 100 \times \frac{\sum \text{numerator}}{\sum \text{denominator}}$$

**Rule 2 — Aggregate as totals, never as a mean of percentages.** A region's
value is the summed numerators over the summed denominators across its woredas.
Averaging woreda percentages would let a tiny woreda count as much as a large
one. (This is also the EPHI house rule for any Total row.)

**Rule 3 — Complete pairs only, and a proportion's numerator cannot exceed its
denominator.** A woreda contributes to a ratio only when it has both a numerator
and a denominator greater than zero, and for a proportion the numerator must be
less than or equal to the denominator. This single rule is what keeps every
percentage at or below 100%. It is the fix for the >100% values seen in an early
draft (a timeliness figure of 122% came from rows with a numerator but no
denominator, and from a leaked "total" row of 15/3).

**Rule 4 — "Not reported" is not zero.** A blank cell is `NA` (not reported); a
typed `0` is a genuine zero. We never let `na.rm` quietly turn missing data into
a zero, because a region can look like 0% purely because nobody filled the cell
in. Every output therefore also carries a *reporting rate*, so performance and
reporting are always read together.

## 1.4 How messy columns become structured records

A workbook's data sheet is a wide strip of repeating
`[Numerator | Denominator | Value]` groups, with no machine-readable grouping.
The pipeline reconstructs the grouping by reading left to right and remembering
the most recent numerator and denominator, then attaching them to the next value
column it meets. Each value column is matched to a canonical indicator by a
**keyword in its header**, not by its position, so a region reordering its
columns does not break anything.

## 1.5 How a reported name becomes a map

Reported woreda names never match the 2021 shapefile directly. We resolve them
through a chain of the cleaning artefacts you maintain:

```
reported name ─► Gname (cleaning key) ─► reference woreda (region-scoped) ─► ADM3_PCODE
```

Scoping the candidates to the reporting region is what removes ambiguity when a
woreda name occurs in more than one region. Joining on `ADM3_PCODE` (never on
names) is what lets the data meet the shapefile cleanly.

## 1.6 Two report products from one dataset

- **`kpi_report.qmd`** is the standing single-period report. It is pinned to
  `FOCUS_PERIOD`, which auto-advances to the newest quarter, so it always
  describes the latest period on its own.
- **`kpi_trends.qmd`** is the cross-period report. It discovers every period in
  the data and compares them, with the data-quality story foregrounded.

When a new quarter arrives the single-period report becomes that quarter's report
and the previous render is archived under its period name, so prior reports are
preserved.

---

# Part II — The code, walked through with examples

```
raw xlsx/docx ─► convert.R ─► converted/  (readable copies, optional)

                 catalog.R   (config: 23 indicators + per-period read rules)
                     │ sourced by
                     ▼
master ◄─ harmonize.R ─► output/master_woreda.csv  (woreda × indicator × PERIOD)
                          output/kpi_region.csv, kpi_national.csv

names ─► crosswalk_build.R ─► output/crosswalk_woreda.csv   (name → ADM3_PCODE)
            uses: All GIS names.xlsx + reference_geo_names.csv

geocode ◄─ spatialize.R ─► output/master_woreda_geo.csv  (+ population backfill)
                           output/kpi_woreda_pcode.csv
                           uses: Population by GIS woreda best.xlsx

figures ◄─ analyze.R + analyze_depth.R  (FOCUS_PERIOD)  ─► exports/figures, exports/tables
           analyze_trends.R             (ALL periods)   ─► exports/figures/T0*, trend_*.csv

reports ◄─ kpi_report.qmd, kpi_trends.qmd, pipeline_documentation.qmd ─► HTML + Word
```

Run everything from the project root:

```r
Rscript scripts/run_all.R
```

`run_all.R` calls each script in order with `system2("Rscript", …)`, then renders
the reports and archives the single-period report under its period. Each script
is independent and re-runnable; they communicate only through CSVs in `output/`.

## 2.1 `catalog.R` — the configuration you edit most

Two objects, both plain tibbles.

**`KPI_CATALOG`** defines the 23 indicators. Each row gives a stable `code`, the
PHEM `group`, a `value_type` (`proportion` or `rate`), a `label`, and a
**`pattern`** — a regex matched against the value-column header in the workbooks.

```r
  4, "RPT_TIMELY", "Reporting", "proportion",
     "Surveillance reports submitted on time (timeliness)",
     "submitted surveillance reports(?!.*complet)(?!.*through dhis)"
```

The regex matters because columns move. "Proportion of submitted surveillance
reports…" is a far more stable handle than "column 11". The negative lookaheads
stop the timeliness pattern from also catching the completeness and DHIS2
columns, which share the opening words.

**`REGION_CONFIG`** is now **multi-period**. There is one tribble block per
period, each row describing how to read one workbook (`file`, `sheet`,
`header_row`, `region`), and the block stamps its own `dir` and `period`:

```r
cfg_wk2852 <- tibble::tribble(
  ~file, ~sheet, ~header_row, ~region, ~notes,
  "DDAHB, week 28-52, 2025 KPI report.xlsx", "data collection tool", 2L, "Dire Dawa", "title row offset; facility level",
  ...
) |> dplyr::mutate(dir = "raw/2025-Wk28-52", period = "2025-Wk28-52")

REGION_CONFIG <- dplyr::bind_rows(cfg_wk0315, cfg_wk1627, cfg_wk2852)
```

`harmonize.R` loops over **every row of every block**, so all periods are read in
one pass. `header_row = 2L` is how the Dire Dawa and Central Ethiopia title rows
are skipped. Three more values steer the rest of the pipeline:

```r
PERIOD_LEVELS <- c("2025-Wk03-15", "2025-Wk16-27", "2025-Wk28-52")  # oldest -> newest
PERIOD_LABELS <- c("2025-Wk28-52" = "Q3-Q4 · Wk 28-52", ...)        # friendly axis labels
FOCUS_PERIOD  <- tail(PERIOD_LEVELS, 1)                             # newest = single-period report
```

```{r}
read_csv(op("indicator_catalog.csv"), show_col_types = FALSE) |>
  select(order, code, group, value_type, target) |> head(6) |>
  kable(caption = "First rows of the catalog (output/indicator_catalog.csv).")
```

## 2.2 `harmonize.R` — building the master

### Step 1: classify every column

`classify()` tags each header from its leading words:

```r
if (str_detect(hl, "^proportion|attack rate|case-?fatality rate")) return("value")
if (str_detect(hl, "^total number|^number of expected"))           return("denom")
if (str_detect(hl, "^number of|^district +for which|^total"))      return("num")
# plus region / zone / woreda / date / population / sn
```

### Step 2: the left-to-right walk (worked example)

Imagine a five-column slice of a sheet, already classified:

```
header:   Number of reports on time | Total reports expected | Proportion timely | Number of responses | Proportion responded
role:     num                       | denom                  | value             | num                 | value
```

The walk remembers the latest `num` and `denom`, and emits a triplet at each
`value`:

```r
last_num <- NA; last_denom <- NA
for (j in start:length(hdr)) {
  if      (roles[j] == "num")   last_num   <- j
  else if (roles[j] == "denom") last_denom <- j
  else if (roles[j] == "value")
    triplets[[length(triplets)+1]] <- list(value_col = j, num_col = last_num,
                                            denom_col = last_denom, header = hdr[j])
}
```

Result: "Proportion timely" pairs with columns 1 and 2; "Proportion responded"
pairs with column 4 and *reuses* denominator column 2 (the nearest preceding
denominator). Rates are the exception — they carry two "Number of…" columns, so
after matching we override them positionally (`num = value-2, denom = value-1`).

### Step 3: name each value column

`match_code()` tests a value header against every catalog `pattern` and takes the
first hit. If a region repeats an indicator (a disease-specific copy), the
duplicate is dropped. If a column matches nothing but the count of value columns
equals the catalog length, an order-based fallback fills the gaps.

### Step 4: drop aggregate rows (worked example)

Regions leak "Total" and "Regional cumulative" rows into the woreda list, and
Addis puts a regional-total line under the sub-city it reports by. `is_aggregate()`
catches both the unit cell and the zone cell:

```r
is_aggregate <- function(x) {
  xl <- str_squish(tolower(ifelse(is.na(x), "", x)))
  xl %in% c("total","regional cumulative","region","grand total","sum",
            "cumulative","aa","a.a","addis ababa") |
    str_detect(xl, "cumu|grand total|^total$|^sum$|^region$")
}
keep <- !blank_woreda & !is_aggregate(geo$woreda) & !is_aggregate(geo$zone)
```

The `cumu` fragment is deliberate: it catches both "cumulative" and the misspelt
"cumuative" that appeared in one file. Without this, a single leaked total row
would double-count an entire region.

### Step 5: the four rules in code

```r
value_pct  = 100 * numerator / denominator,                    # Rule 1
reported   = !is.na(numerator) | !is.na(denominator) | !is.na(value_reported), # Rule 4
valid_pair = !is.na(numerator) & !is.na(denominator) & denominator > 0 &
             (value_type == "rate" | numerator <= denominator) # Rule 3
```

and the rollup applies Rules 1 and 2 together over complete pairs only:

```r
summarise(num_sum   = sum(numerator[valid_pair]),    # complete pairs only
          denom_sum = sum(denominator[valid_pair])) |>
  mutate(value_pct = 100 * num_sum / denom_sum)      # totals, not mean of %
```

`roll(master, region, period)` and `roll(master, period)` produce the region and
national tables. Because `period` is in the grouping, **every period is rolled up
independently in the same call**.

```{r}
read_csv(op("kpi_national.csv"), show_col_types = FALSE) |>
  filter(value_type == "proportion") |>
  count(period, name = "indicator_rows") |>
  kable(caption = "kpi_national.csv holds every period (grouped, not mixed).")
```

## 2.3 `crosswalk_build.R` — names to `ADM3_PCODE`

`match_one(region, zone, woreda)` tries, in order: a manual override, the Addis
sub-city rule, the cleaning key, a direct exact match, then a fuzzy match
(`adist`) as a last resort. A worked example of the chain for one name:

```
reported "Wonbera" (Benishangul Gumuz)
  ─► MANUAL_OVERRIDE says Wonbera → "Wembera"
  ─► reference row with woreda "Wembera" AND region "Benishangul Gumz"
  ─► ADM3_PCODE  ET060201
```

Region scoping is the key step. The same woreda name can occur in several
regions, so candidates are filtered to the reporting region (`REGION_XWALK`
translates the KPI region name to the 2021 reference name) before any tie-break.
The crosswalk runs on the distinct names in the master, so it is period-agnostic
— one mapping serves every period. To fix a stubborn name, add a line:

```r
"Gambella", "Gambella Woreda", "Gambela town"   # region, reported, reference name
```

Tigray and Central Ethiopia were added to `REGION_XWALK` (they match the 2021
names exactly). Tigray, reporting for the first time, has ~17 woreda names not
yet in the cleaning key; these are flagged in `crosswalk_unmatched.csv` and do
not affect Tigray's region-level numbers.

## 2.4 `spatialize.R` — geocode, roll to woreda, backfill population

Joins the master to the crosswalk on `(region, zone, woreda)` to attach
`adm3_pcode`, then rolls up to the canonical woreda (P-code) level using the same
complete-pair rule. Two additions are worth noting.

**Population backfill.** Many regions leave Population blank, which blocks
population-based metrics such as attack rate. `Population by GIS woreda best.xlsx`
maps a clean shapefile woreda name to a population. We join it on the crosswalked
reference name and coalesce:

```r
population_best = coalesce(population, population_ref)   # reported first, else reference
```

`population_best` is what an attack-rate or incidence calculation should use. A
compact lookup is written to `output/woreda_population.csv`.

**Wide master.** `master_woreda_wide.csv` pivots indicators into columns (one row
per woreda × period) for quick browsing in the Data Explorer.

## 2.5 `analyze.R` / `analyze_depth.R` — the single-period figures

Both **source `catalog.R` and filter to `FOCUS_PERIOD`**, so the standing report
shows one period:

```r
region   <- filter(region,   period == FOCUS_PERIOD)
national <- filter(national, period == FOCUS_PERIOD)
```

Conventions: the performance palette is red→yellow→green (`perf_fill()`), with
grey for not-reported; the grey→orange→red ramp is reserved for count maps. The
heatmap appends a National column; the region map greys non-reporting regions;
the woreda map joins `kpi_woreda_pcode` to ADM3 polygons on `adm3_pcode`.

## 2.6 `analyze_trends.R` — the cross-period figures

Reads the all-period rollups, **discovers the periods from the data**, orders
them by `PERIOD_LEVELS`, and draws the newest period bold and red (house style).
It emits the participation matrix (T01), coverage trend (T02), data-quality
scorecard (T03), reporting-completeness trend (T04), national trajectory (T05),
core-function bars (T06), and region composite trend (T07), plus `trend_*.csv`
for each.

It also carries an explicit guardrail. Although Rule 3 already keeps every
rollup at or below 100%, the script asserts it and stops the run if any future
data slips through:

```r
bad <- bind_rows(region, national) |>
  filter(value_type == "proportion", value_pct > 100.0001)
if (nrow(bad) > 0) stop("Guardrail tripped: a proportion exceeds 100%. ...")
```

## 2.7 Output data model

```{r}
tibble::tribble(
  ~File, ~Grain, ~Key,
  "master_woreda.csv", "woreda × indicator × period (long)", "region, period, woreda, indicator",
  "master_woreda_geo.csv", "as above + geocode + population_best", "+ adm3_pcode",
  "master_woreda_wide.csv", "woreda × period, indicators as columns", "region, woreda, period",
  "kpi_region.csv", "region × indicator × period", "region, period, indicator",
  "kpi_national.csv", "national × indicator × period", "period, indicator",
  "kpi_woreda_pcode.csv", "woreda(P-code) × indicator × period", "adm3_pcode, period, indicator",
  "woreda_population.csv", "woreda population lookup", "adm3_pcode",
  "crosswalk_woreda.csv", "reported name → P-code", "region, zone, woreda",
  "trend_*.csv", "cross-period comparison tables", "period (+ region/indicator)"
) |> kable()
```

---

# Part III — Reproducing this next quarter

This is the workflow to run when the next quarter's workbooks arrive. Nothing
here requires rewriting logic — only configuration and a re-run.

**1. Place the files.** Create `raw/<new-period>/` (one folder per period, e.g.
`raw/2025-Wk53-65/`) and drop that quarter's workbooks in. Keep the shared,
period-independent files in the project root: `All GIS names.xlsx`,
`reference_geo_names.csv`, and `Population by GIS woreda best.xlsx`.

**2. Inspect the new workbooks once.** Sheet names and header rows drift between
quarters. From R:

```r
source("scripts/view.R")
kpi_files()                 # lists every file under raw/ (all periods)
sheets("Tigray")            # which sheet holds the data
vx("Tigray", "Woreda")      # open it in the Data Explorer to find the header row
```

Note, for each file, the data sheet name and the 1-based header row (the row
holding "Region or city administration", "Proportion of…", and so on).

**3. Add one config block** in `scripts/catalog.R`: copy a `cfg_<period>` tribble,
fill in `file` / `sheet` / `header_row` / `region` for each workbook, append
`|> dplyr::mutate(dir = "raw/<new-period>", period = "<new-period>")`, add it to
the `bind_rows(...)` call, and append the new label to `PERIOD_LEVELS` and
`PERIOD_LABELS`.

**4. Re-run** `Rscript scripts/run_all.R`.

What happens automatically:

- the master gains the new period; all rollups and the wide table extend;
- `FOCUS_PERIOD` advances, so `kpi_report.qmd` becomes the new quarter's standing
  report, and the previous render is archived as
  `reports/archive/kpi_report_<previous-period>.{html,docx}`;
- `kpi_trends.qmd` discovers the new period and extends every figure and table;
- the >100% guardrail and the not-reported / complete-pair rules apply unchanged.

**5. Check the new period parsed correctly.** Read `output/harmonize_log.txt`
(rows kept and value columns found per workbook) and the crosswalk summary
(mapped vs needs-review). Add any new unmatched woreda names to `MANUAL_OVERRIDE`
in `crosswalk_build.R` and re-run.

## Common tasks

| You want to… | Do this |
|---|---|
| Add the next quarter | Part III above |
| Add a region to an existing period | add a row to that period's `cfg_*` block in `catalog.R` |
| Add / rename an indicator | add a row to `KPI_CATALOG` with a distinctive `pattern` |
| Set official targets | edit the `KPI_CATALOG$target` line in `catalog.R` |
| Fix an unmatched woreda | add a row to `MANUAL_OVERRIDE` in `crosswalk_build.R` |
| Change which period the standing report shows | it tracks the newest period; to pin another, set `FOCUS_PERIOD` in `catalog.R` |
| Rebuild everything | `Rscript scripts/run_all.R` |

## Caveats

- **Targets are an 80% benchmark**, editable in `catalog.R`.
- **Addis granularity differs by period** — sub-city level in Wk 3–15, woreda
  level in the other two. Proportions stay comparable; woreda counts do not.
- **Period length differs** — Wk 28–52 is two quarters, the others one each.
  Proportions are comparable; raw counts scale with length.
- **Dire Dawa** reports by facility; most facility names are not in the cleaning
  key, so its within-region woreda map is incomplete (region numbers are fine).
- **Tigray** woreda names are ~80% mapped (new region, cleaning key incomplete).
- **Disease-specific attack-rate / CFR** columns are dropped; the generic rate is
  kept. Splitting them, and computing attack rate from `population_best`, are the
  natural next enhancements.
